---
title: "Week 06: R for Data Analytics (Data Screening Part 1)"
author: "Ziyuan Huang"
date: "Last Updated: `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  fig.width = 8,
  fig.height = 5
)

# Reproducible seeds using seedhash (same pattern as Weeks 02-05)
# install.packages("seedhash") # Uncomment if not installed
library(seedhash)
gen <- SeedHashGenerator$new("Week06 Data Screening - ANLY500")
seeds <- gen$generate_seeds(10)
print(seeds)
set.seed(seeds[1])

cat("MD5 Hash:", gen$get_hash(), "\n")
cat("Using seed:", seeds[1], "\n")
```

# Introduction

## 1. Overview

This Week 06 tutorial builds on our previous data analytics foundations and focuses on the critical first step of any analysis: **Data Screening**. Before running statistical models, we must ensure our data is clean, accurate, and meets necessary assumptions. This tutorial upgrades the lecture concepts from `06_datascreen_1.rmd` into hands-on, code-first workflows:

-   **Accuracy Checks**: Identifying and fixing typos, out-of-range values, and coding errors.
-   **Missing Data**: Understanding types of missingness (MCAR, MNAR) and handling them appropriately (deletion vs. imputation).
-   **Outlier Detection**: Using univariate (Z-scores) and multivariate (Mahalanobis distance) methods.
-   **Assumptions**: A brief introduction to what we prepare for (Normality, Linearity, etc.).

> **Reference:** This tutorial draws heavily from Field, Miles, and Field [1], *Discovering Statistics Using R*, particularly the chapters on data screening and cleaning, with IEEE-style citations throughout.

> **Upgrade note:** This hands-on tutorial extends the Week 06 lecture slides in `06_datascreen_1.rmd` with code-first workflows, beginner-friendly explanations, and theoretical grounding from Field et al. [1].

### For Beginners: Quick Start

-   **How to run code:**
    -   RStudio: place cursor in a chunk, press Ctrl+Enter (Cmd+Enter on Mac) to run; use Knit to render the full document.
    -   VS Code: highlight lines and press Ctrl+Enter, or click the Run button above a chunk.
-   **Installing packages:** if you see "there is no package called 'X'", run `install.packages("X")` once, then `library(X)`.
-   **If stuck:** restart R (RStudio: Session → Restart R), then re-run chunks from the top (setup first).
-   **Reading errors:** focus on the last lines of the message; they usually state the actual problem (e.g., object not found, package missing, misspelled column).

## 2. Learning Objectives

By the end of this tutorial, you will be able to:

1.  Perform **accuracy checks** on categorical and continuous variables.
2.  Diagnose types of **missing data** (MCAR vs. MNAR).
3.  Implement strategies for **handling missing data**, including listwise deletion and multiple imputation with `mice`.
4.  Detect and handle **univariate outliers** using Z-scores (cutoff $\pm 3.29$).
5.  Detect and handle **multivariate outliers** using Mahalanobis distance.

## 3. Required Packages

```{r load-packages}
# Install if needed
# install.packages(c("tidyverse", "rio", "mice", "VIM", "knitr", "kableExtra", "ggplot2", "gridExtra"))

library(tidyverse)
library(ggplot2)  # For data visualization
library(gridExtra) # For arranging multiple plots
library(rio)      # For data import
library(mice)     # For multiple imputation
library(VIM)      # For visualizing missing data
library(knitr)
library(kableExtra)

cat("\n=== REPRODUCIBLE SEED INFORMATION ===")
cat("\nGenerator Name: Week06 Data Screening - ANLY500")
cat("\nMD5 Hash:", gen$get_hash())
cat("\nAvailable Seeds:", paste(seeds[1:5], collapse = ", "), "...\n\n")
```

---

# Part 1: Data Screening Principles

## 1.1 What is Data Screening?

Data screening is the process of inspecting a dataset for errors, missing values, and outliers before analysis. As Field et al. note, screening is crucial because issues like outliers can "bias the mean and inflate the standard deviation" [1], and missing data can arise from various sources (e.g., participants accidentally missing questions) [1]. Field explicitly refers to Tabachnick and Fidell as the "definitive guide to screening data" [1].

## 1.2 The "Garbage In, Garbage Out" Rule

Data screening is a key facet of any analysis. If the data is incorrect, the output will be useless. Screening ensures results are interpretable and valid. Our checks should always include:

1.  **Accuracy**: Is the data correct? (e.g., "Age" shouldn't be 200).
2.  **Missing Data**: Is the data complete? (e.g., Did someone skip questions?).
3.  **Outliers**: Are there extreme values? (e.g., A millionaire in a study of student income).
4.  **Assumptions**: Does the data fit the rules of our statistical test? (e.g., Is it a bell curve?).

### An Important Note on Significance

Normally, our $\alpha$ value is set to 0.05 (Type 1 Error), meaning we use $p < .05$ as a criterion for statistical significance. However, **in data screening, we want to use a stricter criterion** to avoid over-cleaning or incorrectly flagging data. Therefore, we use **$p < .001$** to denote problematic issues [1].

## 1.3 The Order of Operations

We follow this specific order because each step can affect the next:

1.  **Accuracy Check**: If you correct errors, they may become missing data points.
2.  **Missing Data Check**: If you replace missing data, they may become outliers.
3.  **Outliers Check**: If you exclude outliers, this may change assumption interpretations (e.g., normality).
4.  **Assumptions**: Finally, check assumptions for the specific analysis to be performed.

---

# Part 2: Accuracy Checks

We will use a dataset that examines the resiliency (RS columns) of teenagers after a natural disaster. While we haven't covered specific analyses yet, we can screen this data overall.

```{r import-data}
# Import the dataset
# Note: Ensure "data/data_screening.csv" exists in your working directory or adjust the path
# If the file is not available, we will simulate a similar structure for this tutorial
if(file.exists("data/data_screening.csv")) {
  master <- import("data/data_screening.csv")
} else {
  # Simulation for tutorial purposes if file is missing
  set.seed(seeds[2])
  n <- 100
  master <- data.frame(
    Sex = sample(c(1, 2, 3), n, replace = TRUE, prob = c(0.45, 0.45, 0.1)), # 3 is typo
    SES = sample(c(1, 2, 3), n, replace = TRUE),
    Grade = sample(8:14, n, replace = TRUE), # >12 is typo
    Absences = sample(0:20, n, replace = TRUE), # >15 is typo
    RS1 = sample(1:7, n, replace = TRUE),
    RS2 = sample(1:7, n, replace = TRUE),
    RS3 = sample(c(1:7, 99), n, replace = TRUE, prob = c(rep(0.14, 7), 0.02)) # 99 is outlier/typo
  )
  # Introduce some missingness
  master[sample(1:n, 10), "RS1"] <- NA
  master[sample(1:n, 5), "SES"] <- NA
}

str(master)
```

## 2.1 Categorical Variables

We need to check for factor variables incorrectly imported as numbers and out-of-range values. Let's check `Sex` and `SES`.

### Visualizing the Problem

Before fixing, let's visualize the `Sex` variable to see the error. We expect only two categories (e.g., 1 and 2), but a third bar might appear.

```{r plot-accuracy-cat}
p1 <- ggplot(master, aes(x = factor(Sex))) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Before Cleaning: Sex",
       subtitle = "Note the invalid category '3'",
       x = "Sex Code",
       y = "Count") +
  theme_minimal()

p1
```

**Interpretation:** You can clearly see a bar for "3". If our codebook says 1=Women and 2=Men, then "3" is a typo! This visual confirmation is faster than reading raw tables.

### Checking Tables

```{r accuracy-cat}
notypos <- master # Create a copy to update
apply(notypos[ , c("Sex", "SES")], 2, table)
```

**Explanation:**
-   `apply(X, 2, table)`: This function takes our data (`X`), looks at columns (`2`), and runs the `table()` function on each one.
-   It's a quick way to get counts for multiple columns at once.

### Fixing the Issue

We can fix this by using `factor()` and excluding the bad label. Values not listed in `levels` will automatically become `NA`.

```{r fix-cat}
## Fix categorical labels and drop typos
notypos$Sex <- factor(notypos$Sex, 
                     levels = c(1,2), # We intentionally omit 3
                     labels = c("Women", "Men"))

notypos$SES <- factor(notypos$SES, 
                     levels = c(1,2, 3),
                     labels = c("Low", "Medium", "High"))

# Let's verify with a plot
p2 <- ggplot(notypos, aes(x = Sex)) +
  geom_bar(fill = "forestgreen", color = "black") +
  labs(title = "After Cleaning: Sex",
       subtitle = "Invalid categories removed (became NA)",
       x = "Sex",
       y = "Count") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

**Visual Check:** The "Before" plot shows the error (3). The "After" plot shows only valid categories (Women, Men). The typos didn't disappear; they became `NA` (missing values), which we will deal with later.

## 2.2 Continuous Variables

Use `summary()` to check min/max values. For example, if the RS scale runs from 1 to 7, any value outside this range (like 99) is a typo.

```{r accuracy-cont}
summary(notypos)
```

### Visualizing Continuous Errors

Let's look at `RS3`. We suspect a typo (99).

```{r plot-accuracy-cont}
# Histogram to see distribution shape
p3 <- ggplot(notypos, aes(x = RS3)) +
  geom_histogram(fill = "orange", color = "black", bins = 30) +
  labs(title = "Histogram of RS3",
       subtitle = "Note the gap and value near 100",
       x = "RS3 Score", y = "Frequency") +
  theme_minimal()

# Boxplot to see outliers clearly
p4 <- ggplot(notypos, aes(y = RS3)) +
  geom_boxplot(fill = "orange", alpha = 0.5) +
  labs(title = "Boxplot of RS3",
       subtitle = "The dot at top is the extreme value",
       y = "RS3 Score") +
  theme_minimal()

grid.arrange(p3, p4, ncol = 2)
```

**Interpretation:**
-   **Histogram (Left):** Most data is clustered 0-10. There is a tiny bar near 100. This is suspicious.
-   **Boxplot (Right):** The box is "squashed" at the bottom, and a single dot is way up at 100. This confirms we have an out-of-range value that needs fixing.

### Fixing Issues

How do we fix these?
1.  Find the original data and figure out what the point should be.
2.  **Delete that data point** (set to `NA`) but not the entire participant.

**Example: Using Logical Subsetting**
Suppose `Grade` cannot be > 12 and `Absences` cannot be > 15.

```{r fix-cont-specific}
# Check Grade
summary(notypos$Grade)
# Code Explanation: "In the Grade column, find values > 12, and set them to NA"
notypos$Grade[ notypos$Grade > 12 ] <- NA 
summary(notypos$Grade)

# Check Absences
summary(notypos$Absences)
notypos$Absences[ notypos$Absences > 15 ] <- NA
summary(notypos$Absences)
```

**Example: Multiple Columns**
If multiple columns (e.g., RS items) have the same range (1-7), we can clean them all at once.

```{r fix-cont-multi}
# Assuming RS columns are from column 5 onwards (adjust indices as needed)
# In our simulated data, RS columns are 5, 6, 7
rs_cols <- grep("RS", names(notypos))

# Check before
summary(notypos[ , rs_cols])

# Apply fix: any value > 7 in these columns becomes NA
notypos[ , rs_cols][ notypos[ , rs_cols] > 7 ] <- NA

# Check after
summary(notypos[ , rs_cols])
```

---

# Part 3: Missing Data

Use `summary()` or `apply()` to count `NA` values.

```{r missing-check}
apply(notypos, 2, function(x) { sum(is.na(x)) })
```

## 3.1 Diagnosing Missingness

Understanding *why* data is missing is crucial [1]:

1.  **MCAR (Missing Completely at Random):**
    -   **Analogy:** You are carrying a tray of blood samples and trip, breaking random vials. The missingness has nothing to do with the patient's health.
    -   **Handling:** Safe to replace or ignore.
2.  **MNAR (Missing Not at Random):**
    -   **Analogy:** Patients with very high blood pressure refuse to have it measured because they are scared. The missingness *is* related to the health outcome.
    -   **Handling:** **Dangerous** to replace. Excluding might be safer.

> **Quick Fix:** For simple statistics, you can often use `na.rm = TRUE` (e.g., `mean(x, na.rm = TRUE)`) to ignore missing values without permanently deleting them [1].

## 3.2 Visualizing Missing Data

The `VIM` package helps visualize patterns.

```{r vim-plot}
aggr(notypos, numbers = TRUE, prop = FALSE)
```

**Interpretation:**
-   **Left Bar Chart:** Shows the proportion of missing data in each variable.
-   **Right Grid:** Shows the *combinations* of missingness. Blue is observed, Red is missing.
    -   If you see red blocks lining up across multiple variables, it means people who missed Question A also tended to miss Question B.

## 3.3 Handling Missing Data

### By Rows (Participants)

Check the percentage of missing data per participant.

```{r missing-rows}
# Create a custom function to calculate % missing
percentmiss <- function(x){ sum(is.na(x))/length(x) * 100 }

# Apply it to rows (margin = 1)
missing <- apply(notypos, 1, percentmiss)
table(missing)
```

We can create subsets:
-   **Replaceable:** Participants with $\le 5\%$ missing data.
-   **Not Replaceable:** Participants with $> 5\%$ missing data (often excluded or analyzed separately).

```{r subset-rows}
replace_rows <- subset(notypos, missing <= 5)
noreplace_rows <- subset(notypos, missing > 5)

cat("Original N:", nrow(notypos), "\n")
cat("Rows to replace:", nrow(replace_rows), "\n")
cat("Rows dropped/kept separate:", nrow(noreplace_rows), "\n")
```

### Multiple Imputation with `mice`

We do not replace missing data for categorical/demographic variables (Sex, SES, Grade). We use them to help estimate missing values in other variables.

**What is MICE?**
*Multivariate Imputation by Chained Equations*. It creates multiple "complete" datasets by guessing missing values based on relationships with other variables. It's better than just plugging in the mean because it preserves the variance in the data.

```{r mice-imputation}
# Run mice on our replaceable rows
# printFlag = FALSE keeps the output clean
temp_no_miss <- mice(replace_rows, printFlag = FALSE, seed = 123)
```

Now, complete the dataset (pick one of the imputations, e.g., #1).

```{r mice-complete}
nomiss <- complete(temp_no_miss, 1)
all_columns <- nomiss # This is our clean dataset
```

### Visual Verification: Did Imputation Change Data Shape?

It is critical to verify that imputing data didn't distort our variables. The distribution should look roughly the same before and after.

```{r plot-imputation-check}
# Combine data for plotting
before <- data.frame(Value = replace_rows$RS1, Type = "Original (with NA)")
after <- data.frame(Value = all_columns$RS1, Type = "Imputed (Complete)")
comparison <- rbind(before, after)

ggplot(comparison, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot: RS1 Before vs. After Imputation",
       subtitle = "Curves should be similar. Drastic changes indicate bad imputation.",
       x = "RS1 Score",
       y = "Density") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red"))
```

**Interpretation:** If the Red and Blue curves overlap significantly, your imputation is good! It means you filled in the holes without changing the "story" of the data.

---

# Part 4: Outliers

**Definition:** A case with an extreme value on one or multiple variables.
**Why remove them?** The mean is sensitive to outliers. We use a strict criterion ($p < .001$) to remove only very extreme scores.

## 4.1 Univariate Outliers (Z-scores)

For a single variable, we can use **Z-scores**.
-   **Concept:** A Z-score tells you how many standard deviations a score is from the mean.
    -   $Z = 0$: Exactly average.
    -   $Z = +1$: One standard deviation above average.
    -   $Z = +3.29$: Extremely high (less than 0.1% chance).
-   **Criterion:** $Z > 3.29$ or $Z < -3.29$ (corresponding to $p < .001$) [1].

```{r z-scores}
# Calculate Z-scores for RS1 (as an example)
# scale() function calculates Z-scores
all_columns$RS1_z <- scale(all_columns$RS1)

# Check for outliers
summary(all_columns$RS1_z)
```

### Visualizing Z-scores

```{r plot-zscores}
ggplot(all_columns, aes(x = RS1_z)) +
  geom_histogram(bins = 20, fill = "purple", color = "black") +
  geom_vline(xintercept = c(-3.29, 3.29), linetype = "dashed", color = "red", size = 1) +
  labs(title = "Histogram of Z-scores (RS1)",
       subtitle = "Red dashed lines indicate the +/- 3.29 cutoff",
       x = "Z-score",
       y = "Count") +
  theme_minimal()
```

**Interpretation:** Any bars *outside* the red dashed lines represent outliers that should be removed or investigated.

## 4.2 Multivariate Outliers: Mahalanobis Distance

For multiple continuous variables, we measure the distance of a case from the centroid (mean of means).

**Concept:** Imagine a cloud of points in 3D space. The "centroid" is the middle of the cloud. Mahalanobis distance measures how far a point is from this middle, accounting for the shape of the cloud.

### Calculation

```{r mahalanobis}
# Select only continuous variables for outlier check
# Let's use RS1, RS2, RS3, and Absences
cont_vars <- all_columns %>% select(RS1, RS2, RS3, Absences)

# Calculate distance
mahal <- mahalanobis(cont_vars,
                    colMeans(cont_vars, na.rm=TRUE),
                    cov(cont_vars, use ="pairwise.complete.obs"))
```

### Cutoff and Elimination

-   **Cutoff:** $\chi^2$ (Chi-square) distribution with $df =$ number of variables, at $\alpha < .001$.

```{r outlier-cutoff}
cutoff <- qchisq(1 - .001, ncol(cont_vars))

cat("Degrees of Freedom:", ncol(cont_vars), "\n")
cat("Cutoff Value:", cutoff, "\n")

# Check how many outliers
# FALSE means NOT an outlier (Good), TRUE means outlier (Bad)
summary(mahal < cutoff)
```

## 4.3 Visualizing Outliers

Let's plot the Mahalanobis distances to see who the "troublemakers" are.

```{r plot-mahal}
# Create a data frame for plotting
outlier_plot <- data.frame(
  ID = 1:length(mahal),
  Distance = mahal,
  Outlier = mahal > cutoff
)

ggplot(outlier_plot, aes(x = ID, y = Distance, color = Outlier)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_hline(yintercept = cutoff, linetype = "dashed", color = "red") +
  geom_text(aes(label = ifelse(Outlier, ID, "")), vjust = -0.5, show.legend = FALSE) +
  labs(title = "Mahalanobis Distances",
       subtitle = paste("Points above the red line (Cutoff =", round(cutoff, 2), ") are multivariate outliers"),
       y = "Distance (D^2)",
       x = "Participant ID") +
  theme_minimal() +
  scale_color_manual(values = c("black", "red"))
```

**Interpretation:**
-   Points **below the red line** are safe.
-   Points **above the red line** are multivariate outliers. We've labeled their IDs so you can find them!

### Final Removal

```{r remove-outliers}
# Eliminate outliers
noout <- subset(all_columns, mahal < cutoff)

cat("\nDimensions before outlier removal:", dim(all_columns)[1], "\n")
cat("Dimensions after outlier removal:", dim(noout)[1], "\n")
```

# Summary

In this tutorial, we have covered the essential first steps of data analysis:
1.  **Accuracy:** Checked for and fixed typos and impossible values.
2.  **Missing Data:** Identified missingness patterns and used multiple imputation.
3.  **Outliers:** Applied a strict statistical criterion ($p < .001$) to identify univariate and multivariate outliers.

Your dataset `noout` is now screened and ready for assumption checking and analysis!

# References

[1] A. Field, J. Miles, and Z. Field, *Discovering Statistics Using R*. London: SAGE Publications, 2012.
