---
title: "Week 01: Introduction to R"
author: "Ziyuan Huang"
date: "Last Updated: `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  fig.width = 8,
  fig.height = 5
)
```

```{r machine-info, echo=FALSE, results='asis'}
cat("**Machine:**", paste0(Sys.info()['nodename'], " - ", Sys.info()['sysname'], " ", Sys.info()['release']), "\n\n")
```

# Introduction

## 1. Welcome to R!

Welcome to your first week learning R, a powerful programming language for data analysis and statistical computing. This course is designed to help you become comfortable with R even if you've never programmed before. We'll start with the fundamentals and build your skills step by step.

> **Key term – R:** R is both a programming language and an environment for statistical computing and graphics. Originally developed by Ross Ihaka and Robert Gentleman at the University of Auckland, R has become one of the most widely used tools in data analytics (R Core Team, 2024).

## 2. Learning Objectives

By the end of this week, you will be able to:

1. Understand what R is and why it's valuable for data analysis
2. Navigate the RStudio interface confidently
3. Create and work with different types of data (numbers, text, categories)
4. Store and organize data using vectors, matrices, data frames, and lists
5. Perform calculations and comparisons
6. Load data from CSV files
7. Access and subset specific parts of your data
8. Handle missing data appropriately
9. Use built-in R functions for analysis
10. Write your own simple functions

## 3. What is R and Why Use It?

R offers several advantages for data analytics:

- **Free and open-source**: No licensing costs, ever
- **Comprehensive**: Over 19,000 packages for specialized analyses
- **Reproducible**: Your analysis can be easily shared and repeated
- **Visualization**: Publication-quality graphics with ggplot2
- **Community**: Extensive online support and resources
- **Integration**: Works with databases, web APIs, and other languages

> **Key concept – Reproducible Research:** Field, Miles, and Field (2012, ch. 3) emphasize that reproducible research allows others to verify your findings and build upon your work. Using R scripts instead of point-and-click software creates an auditable trail of your analytical decisions.

---

# Part 1: Getting Started with R

## 1.1 Installing R and RStudio

### Installation Steps

1. **Download R** from: https://cran.r-project.org/
   - Choose your operating system (Windows, macOS, or Linux)
   - Install R first

2. **Download RStudio** from: https://posit.co/downloads/
   - RStudio is an Integrated Development Environment (IDE)
   - Makes working with R much easier
   - Install RStudio after R

### The RStudio Interface

RStudio has four main panes:

1. **Source/Script Editor** (top-left): Where you write and save your code
2. **Console** (bottom-left): Where code executes and results appear
3. **Environment/History** (top-right): Shows objects you've created
4. **Files/Plots/Help** (bottom-right): Navigation and visualization

```{r eval=FALSE}
# You can type directly in the Console:
2 + 2

# Or write code in the Script Editor and run it (Ctrl+Enter or Cmd+Return):
x <- 10
print(x)
```

## 1.2 R as a Calculator

R can perform all basic arithmetic operations:

```{r calculator}
# Addition
5 + 3

# Subtraction
10 - 4

# Multiplication
6 * 7

# Division
20 / 4

# Exponentiation
2^8

# Modulo (remainder)
17 %% 5

# Order of operations (PEMDAS applies)
(5 + 3) * 2^2
```

> **Key concept – Order of Operations:** Like in mathematics, R follows PEMDAS (Parentheses, Exponents, Multiplication/Division, Addition/Subtraction). Use parentheses to control calculation order (Field et al., 2012, ch. 3).

## 1.3 Creating and Managing Objects

In R, we store values in **objects** using the assignment operator `<-`:

```{r objects}
# Create objects
A <- 5
B <- 10
name <- "R Programming"

# Use objects in calculations
A + B
A * B

# Check what's in your workspace
ls()

# View specific object
print(A)
print(name)
```

### Naming Conventions

Follow these rules for naming objects:

- ✅ Use letters, numbers, dots (`.`), and underscores (`_`)
- ✅ Start with a letter (not a number)
- ✅ Be descriptive: `student_age` is better than `x`
- ✅ Use consistent style: `snake_case` or `camelCase`
- ❌ Don't use spaces
- ❌ Don't use special characters (`@`, `#`, `$`, etc.)
- ⚠️ R is case-sensitive: `data` and `Data` are different!

```{r naming-examples}
# Good names
student_age <- 20
total_score <- 95
mean_temperature <- 72.5

# Avoid these
x <- 20              # Not descriptive
my.variable <- 10    # Dots are discouraged in modern R
2ndScore <- 85       # Can't start with number (ERROR!)
```

### Removing Objects

```{r removing}
# Remove a specific object
temp_value <- 100
rm(temp_value)

# Remove multiple objects
x <- 1
y <- 2
z <- 3
rm(x, y, z)

# Remove all objects (use with caution!)
# rm(list = ls())
```

---

# Part 2: Data Types in R

## 2.1 Basic Data Types

R recognizes several fundamental data types:

```{r data-types}
# Numeric (numbers)
age <- 25
height <- 175.5
class(age)          # Check data type
class(height)

# Character (text)
name <- "John Doe"
species <- "Adelie"
class(name)

# Logical (TRUE/FALSE)
is_student <- TRUE
has_passed <- FALSE
class(is_student)

# Integer (whole numbers)
count <- 42L        # The 'L' specifies integer
class(count)
```

### Type Conversion

You can convert between types using `as.*()` functions:

```{r type-conversion}
# Numeric to character
num <- 42
char_num <- as.character(num)
class(char_num)

# Character to numeric
text_num <- "123"
converted <- as.numeric(text_num)
class(converted)

# Numeric to logical
as.logical(1)       # TRUE
as.logical(0)       # FALSE
```

## 2.2 Special Values

R has special values for unusual situations:

```{r special-values}
# NA - Missing/Not Available
missing_value <- NA
is.na(missing_value)

# NULL - Empty/Undefined
empty_value <- NULL
is.null(empty_value)

# Inf - Infinity
positive_inf <- 1/0
negative_inf <- -1/0
print(positive_inf)

# NaN - Not a Number (undefined mathematical operations)
undefined <- 0/0
is.nan(undefined)
```

## 2.3 Factors

Factors represent categorical data:

```{r factors}
# Create a factor
species <- factor(c("Adelie", "Gentoo", "Chinstrap", "Adelie", "Gentoo"))
print(species)

# Check levels
levels(species)

# Specify level order (useful for ordinal data)
satisfaction <- factor(
  c("Low", "High", "Medium", "Low", "High"),
  levels = c("Low", "Medium", "High"),
  ordered = TRUE
)
print(satisfaction)
```

> **Key concept – Factors:** Field, Miles, and Field (2012, ch. 3) explain that factors are essential for categorical variables in R. Unlike character strings, factors store levels, which is memory-efficient and necessary for many statistical models.

---

# Part 3: Data Structures

## 3.1 Vectors

Vectors are collections of elements of the same type:

```{r vectors}
# Numeric vector
ages <- c(23, 25, 31, 28, 26)
print(ages)

# Character vector
names <- c("Alice", "Bob", "Carol", "David", "Eve")
print(names)

# Logical vector
passed <- c(TRUE, TRUE, FALSE, TRUE, FALSE)
print(passed)

# Sequences
seq1 <- 1:10                    # 1 to 10
seq2 <- seq(0, 100, by = 10)    # 0 to 100 by 10
rep1 <- rep(5, times = 3)       # Repeat 5 three times

print(seq1)
print(seq2)
print(rep1)
```

### Vector Operations

```{r vector-operations}
# Element-wise operations
x <- c(1, 2, 3, 4, 5)
y <- c(10, 20, 30, 40, 50)

x + y         # Addition
x * y         # Multiplication
x^2           # Square each element

# Summary statistics
mean(x)
sum(x)
sd(x)         # Standard deviation
length(x)     # Number of elements
```

### Vector Subsetting

```{r vector-subsetting}
# Create a vector
scores <- c(85, 92, 78, 90, 88)

# Access by position
scores[1]           # First element
scores[3]           # Third element
scores[c(1, 3, 5)]  # Multiple elements

# Access by condition
scores[scores > 85]          # Scores greater than 85
scores[scores >= 85 & scores <= 90]  # Between 85 and 90

# Negative indexing (exclude elements)
scores[-2]          # All except second element
```

## 3.2 Matrices

Matrices are 2-dimensional arrays with rows and columns:

```{r matrices}
# Create a matrix
my_matrix <- matrix(
  data = 1:12,
  nrow = 4,
  ncol = 3,
  byrow = FALSE    # Fill by column (default)
)
print(my_matrix)

# Matrix by row
matrix_byrow <- matrix(
  data = 1:12,
  nrow = 4,
  ncol = 3,
  byrow = TRUE
)
print(matrix_byrow)

# Matrix dimensions
dim(my_matrix)
nrow(my_matrix)
ncol(my_matrix)
```

### Matrix Operations

```{r matrix-operations}
# Subsetting matrices
my_matrix[2, 3]      # Row 2, Column 3
my_matrix[1, ]       # First row, all columns
my_matrix[, 2]       # All rows, second column
my_matrix[1:2, 2:3]  # Rows 1-2, Columns 2-3

# Matrix arithmetic
mat1 <- matrix(1:4, nrow = 2)
mat2 <- matrix(5:8, nrow = 2)

mat1 + mat2          # Element-wise addition
mat1 * mat2          # Element-wise multiplication
mat1 %*% mat2        # Matrix multiplication

# Transpose
t(mat1)
```

## 3.3 Data Frames

Data frames are the most common structure for datasets:

```{r dataframes}
# Create a data frame
students <- data.frame(
  Name = c("Alice", "Bob", "Carol", "David"),
  Age = c(20, 22, 21, 23),
  Grade = c(85, 92, 78, 88),
  Passed = c(TRUE, TRUE, TRUE, TRUE)
)

print(students)

# View structure
str(students)

# View dimensions
dim(students)
nrow(students)
ncol(students)

# Column names
names(students)
colnames(students)
```

### Accessing Data Frame Elements

```{r dataframe-access}
# Access columns
students$Name              # Using $
students[, "Grade"]        # Using column name
students[, 3]              # Using position

# Access rows
students[1, ]              # First row
students[c(1, 3), ]        # Rows 1 and 3

# Access specific cells
students[2, 3]             # Row 2, Column 3
students$Grade[2]          # Second value in Grade column
```

### Filtering Data Frames

```{r dataframe-filtering}
# Filter by condition
students[students$Age > 21, ]
students[students$Grade >= 85, ]

# Multiple conditions (AND)
students[students$Age > 20 & students$Grade > 80, ]

# Multiple conditions (OR)
students[students$Age < 21 | students$Grade > 90, ]

# Using subset() function
subset(students, Age > 21)
subset(students, Grade >= 85 & Passed == TRUE)
```

> **Important:** The `subset()` function automatically removes rows with `NA` values. Use bracket notation `[ ]` if you want to keep `NA`s.

## 3.4 Lists

Lists can contain different types and structures:

```{r lists}
# Create a list
my_list <- list(
  numbers = 1:5,
  text = c("a", "b", "c"),
  logical = c(TRUE, FALSE),
  matrix = matrix(1:4, nrow = 2),
  dataframe = students
)

# View structure
str(my_list)

# Access list elements
my_list$numbers              # Using name
my_list[[1]]                 # Using position (double brackets)
my_list[["text"]]            # Using name (double brackets)

# Access nested elements
my_list$dataframe$Name       # Names from dataframe in list
```

---

# Part 4: Working with Data Files

## 4.1 Working Directory

Before importing data, know where R is looking for files:

```{r working-directory}
# Check current working directory
getwd()

# Set working directory (use forward slashes or double backslashes)
# setwd("C:/Users/YourName/Documents/R_Projects")
# setwd("C:\\Users\\YourName\\Documents\\R_Projects")
```

### Best Practices for File Paths

1. **Use R Projects** (recommended): Automatically sets working directory
2. **Use R Markdown**: Automatically uses the file's location
3. **Use relative paths**: `"data/myfile.csv"` instead of full paths
4. **Avoid `setwd()`**: Breaks when you share code

## 4.2 Importing Data

### Using Base R

```{r import-base, eval=FALSE}
# Read CSV files
data_csv <- read.csv("data/mydata.csv")

# Read tab-delimited files
data_tab <- read.table("data/mydata.txt", header = TRUE, sep = "\t")

# Read from clipboard (Windows)
data_clipboard <- read.table("clipboard", header = TRUE)
```

### Using rio Package (Recommended)

The `rio` package handles many file formats automatically:

```{r import-rio, eval=FALSE}
# Install rio if needed
# install.packages("rio")

library(rio)

# Import various formats
data <- import("data/mydata.csv")      # CSV
data <- import("data/mydata.xlsx")     # Excel
data <- import("data/mydata.sav")      # SPSS
data <- import("data/mydata.dta")      # Stata
data <- import("data/mydata.rds")      # R data
```

## 4.3 Exploring Your Data

Once data is loaded, explore it:

```{r explore-data}
# Use built-in airquality dataset
data(airquality)

# View first few rows
head(airquality)

# View last few rows
tail(airquality)

# Structure
str(airquality)

# Summary statistics
summary(airquality)

# Dimensions
dim(airquality)
nrow(airquality)
ncol(airquality)

# Column names
names(airquality)
```

### Interactive Viewing

```{r eval=FALSE}
# Open data viewer (in RStudio)
View(airquality)

# Print specific summaries
summary(airquality$Temp)
table(airquality$Month)
```

---

# Part 5: Basic Operations

## 5.1 Arithmetic Operations

```{r arithmetic}
# Basic operations
10 + 5
10 - 5
10 * 5
10 / 5
10^2

# More complex
(10 + 5) * 2
10 + 5 * 2        # Multiplication before addition

# Functions
sqrt(16)          # Square root
abs(-10)          # Absolute value
log(10)           # Natural log
log10(100)        # Base-10 log
exp(1)            # e^1
```

## 5.2 Comparison Operators

Comparisons return `TRUE` or `FALSE`:

```{r comparison}
# Equality and inequality
5 == 5            # Equal to (two equals signs!)
5 != 3            # Not equal to

# Greater and less than
10 > 5
10 < 5
10 >= 10
10 <= 9

# Compare vectors element-wise
x <- c(1, 2, 3, 4, 5)
x > 3
x == 3
```

## 5.3 Logical Operators

Combine conditions:

```{r logical}
# AND operator (&)
TRUE & TRUE       # TRUE
TRUE & FALSE      # FALSE

# OR operator (|)
TRUE | FALSE      # TRUE
FALSE | FALSE     # FALSE

# NOT operator (!)
!TRUE             # FALSE
!FALSE            # TRUE

# Combining conditions
age <- 25
age > 18 & age < 30              # Between 18 and 30
age < 18 | age > 65              # Less than 18 OR greater than 65
```

### Using Logical Operators for Filtering

```{r logical-filtering}
# Filter airquality data
hot_days <- airquality[airquality$Temp > 85, ]
head(hot_days)

# Multiple conditions
hot_windy <- airquality[airquality$Temp > 85 & airquality$Wind < 10, ]
head(hot_windy)

# Count matching rows
sum(airquality$Temp > 85, na.rm = TRUE)
```

---

# Part 6: Handling Missing Data

## 6.1 Identifying Missing Values

```{r missing-data}
# Create data with missing values
x <- c(1, 2, NA, 4, 5, NA)

# Check for missing values
is.na(x)

# Count missing values
sum(is.na(x))

# Which positions are missing?
which(is.na(x))

# Check airquality for missing values
sum(is.na(airquality))                    # Total NAs
colSums(is.na(airquality))                # NAs per column
```

> **Key concept – Missing Data:** Field, Miles, and Field (2012, ch. 5) discuss three types of missing data mechanisms: Missing Completely At Random (MCAR), Missing At Random (MAR), and Missing Not At Random (MNAR). Understanding why data is missing determines appropriate handling strategies.

## 6.2 Handling Missing Values

```{r handling-missing}
# Remove NAs from vector
x_complete <- x[!is.na(x)]
x_complete

# Or use na.omit()
x_clean <- na.omit(x)
x_clean

# Remove rows with any NA
airquality_complete <- na.omit(airquality)
nrow(airquality)           # Original
nrow(airquality_complete)  # After removing NAs

# Use complete.cases()
complete_rows <- complete.cases(airquality)
airquality_filtered <- airquality[complete_rows, ]
```

## 6.3 Functions and Missing Data

Many functions return `NA` if there are missing values:

```{r functions-missing}
# This returns NA
mean(airquality$Ozone)

# Remove NAs before calculating
mean(airquality$Ozone, na.rm = TRUE)
median(airquality$Ozone, na.rm = TRUE)
sd(airquality$Ozone, na.rm = TRUE)

# Summary handles NAs automatically
summary(airquality$Ozone)
```

---

# Part 7: Essential R Functions

## 7.1 Getting Help

```{r help, eval=FALSE}
# View help documentation
?mean
help(mean)

# Search help
??correlation

# View function arguments
args(mean)

# Run examples
example(mean)
```

## 7.2 Descriptive Statistics

```{r descriptive-stats}
# Create sample data
scores <- c(85, 92, 78, 90, 88, 95, 82, 87)

# Measures of central tendency
mean(scores)        # Average
median(scores)      # Middle value
# mode() doesn't exist in base R

# Measures of dispersion
sd(scores)          # Standard deviation
var(scores)         # Variance
range(scores)       # Min and max
min(scores)
max(scores)

# Other useful functions
sum(scores)
length(scores)
quantile(scores, probs = c(0.25, 0.5, 0.75))  # Quartiles

# Comprehensive summary
summary(scores)
```

## 7.3 Frequency Tables

```{r frequency-tables}
# Create categorical data
grades <- c("A", "B", "A", "C", "B", "A", "B", "A", "C", "B")

# Frequency table
table(grades)

# Proportions
prop.table(table(grades))

# Cross-tabulation (two variables)
data(mtcars)
table(mtcars$cyl, mtcars$gear)

# Add margins (totals)
addmargins(table(mtcars$cyl, mtcars$gear))
```

## 7.4 Data Exploration Functions

```{r exploration-functions}
data(iris)

# Structure
str(iris)

# Summary statistics
summary(iris)

# Dimensions
dim(iris)
nrow(iris)
ncol(iris)

# Column names
names(iris)
colnames(iris)

# First and last rows
head(iris, n = 3)
tail(iris, n = 3)

# Unique values
unique(iris$Species)
length(unique(iris$Species))
```

## 7.5 Correlation and Covariance

```{r correlation}
# Correlation between two variables
cor(iris$Sepal.Length, iris$Sepal.Width)

# Correlation matrix (numeric columns only)
cor(iris[, 1:4])

# Handle missing data
cor(airquality$Ozone, airquality$Temp, use = "complete.obs")

# Covariance
cov(iris$Sepal.Length, iris$Sepal.Width)
cov(iris[, 1:4])
```

---

# Part 8: Writing Your Own Functions

## 8.1 Function Basics

Functions automate repetitive tasks:

```{r function-basics}
# Basic function structure
square <- function(x) {
  result <- x^2
  return(result)
}

# Use the function
square(5)
square(10)

# Simplified version (R returns last value automatically)
square_simple <- function(x) {
  x^2
}

square_simple(7)
```

## 8.2 Functions with Multiple Arguments

```{r function-arguments}
# Function with two arguments
add_numbers <- function(a, b) {
  a + b
}

add_numbers(5, 3)
add_numbers(10, 15)

# Function with default values
greet <- function(name = "Friend") {
  paste("Hello,", name)
}

greet("Alice")
greet()           # Uses default "Friend"
```

## 8.3 Practical Examples

```{r practical-functions}
# Calculate z-score
z_score <- function(x, mean_val, sd_val) {
  (x - mean_val) / sd_val
}

scores <- c(85, 92, 78, 90, 88)
z_score(85, mean(scores), sd(scores))

# Convert Fahrenheit to Celsius
f_to_c <- function(temp_f) {
  temp_c <- (temp_f - 32) * 5/9
  return(temp_c)
}

f_to_c(98.6)
f_to_c(32)
f_to_c(212)

# Function that returns multiple values (using list)
summary_stats <- function(x) {
  list(
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE)
  )
}

summary_stats(scores)
```

---

# Part 9: Combining Data

## 9.1 Combining Vectors and Data Frames

```{r combining-data}
# Combine vectors as columns
x <- c(1, 2, 3)
y <- c(4, 5, 6)
cbind(x, y)

# Combine as rows
rbind(x, y)

# Combine data frames by columns
df1 <- data.frame(ID = 1:3, Name = c("A", "B", "C"))
df2 <- data.frame(Score = c(85, 90, 78))
combined_cols <- cbind(df1, df2)
print(combined_cols)

# Combine data frames by rows (must have same columns)
df3 <- data.frame(ID = 4:5, Name = c("D", "E"), Score = c(92, 88))
combined_rows <- rbind(combined_cols, df3)
print(combined_rows)
```

## 9.2 Adding Columns and Rows

```{r adding-data}
# Add column to data frame
students <- data.frame(
  Name = c("Alice", "Bob", "Carol"),
  Age = c(20, 22, 21)
)

students$Grade <- c(85, 92, 78)
print(students)

# Add row
new_student <- data.frame(Name = "David", Age = 23, Grade = 88)
students <- rbind(students, new_student)
print(students)
```

---

# Part 10: Practical Example - Palmer Penguins

Let's put everything together with a real dataset:

```{r penguins-example}
# Install and load palmerpenguins package
# install.packages("palmerpenguins")
library(palmerpenguins)

# Load the penguins dataset
data(penguins)

# Explore the data
str(penguins)
head(penguins)
summary(penguins)

# How many penguins of each species?
table(penguins$species)

# Average bill length by species
aggregate(bill_length_mm ~ species, data = penguins, FUN = mean, na.rm = TRUE)

# Filter for Adelie penguins only
adelie <- subset(penguins, species == "Adelie")
nrow(adelie)

# Penguins with bill length > 50mm
long_bills <- subset(penguins, bill_length_mm > 50)
nrow(long_bills)

# Create a new variable: bill aspect ratio
penguins$bill_ratio <- penguins$bill_length_mm / penguins$bill_depth_mm
head(penguins$bill_ratio)

# Summary of body mass by sex
aggregate(body_mass_g ~ sex, data = penguins, FUN = mean, na.rm = TRUE)
```

---

# Summary

## Key Takeaways

1. **R Basics**
   - R is free, powerful, and reproducible
   - RStudio makes R easier to use
   - Objects store values with `<-`

2. **Data Types**
   - Numeric, character, logical, factor
   - Special values: NA, NULL, Inf, NaN

3. **Data Structures**
   - Vectors: 1-dimensional, same type
   - Matrices: 2-dimensional, same type
   - Data frames: 2-dimensional, different types
   - Lists: flexible, can contain anything

4. **Working with Data**
   - Import with `read.csv()` or `rio::import()`
   - Explore with `head()`, `str()`, `summary()`
   - Subset with `[ ]`, `$`, or `subset()`

5. **Missing Data**
   - Identify with `is.na()`
   - Remove with `na.omit()` or `na.rm = TRUE`
   - Understand why data is missing (MCAR, MAR, MNAR)

6. **Functions**
   - Use built-in functions for analysis
   - Write your own functions for custom tasks
   - Get help with `?function_name`

## Next Steps

Week 2 will build on these fundamentals:

- Advanced data manipulation with dplyr
- Creating visualizations with ggplot2
- More statistical analyses
- Hypothesis testing basics

## Practice Resources

- **Lab Assignment**: `01_lab_R_learning.Rmd` (30 exercises)
- **RStudio Cheatsheets**: https://posit.co/resources/cheatsheets/
- **Quick-R**: https://www.statmethods.net/
- **Stack Overflow**: https://stackoverflow.com/questions/tagged/r

---

# References

Field, A., Miles, J., & Field, Z. (2012). *Discovering Statistics Using R*. SAGE Publications Ltd., London.

R Core Team (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria. https://www.R-project.org/

Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis* (2nd ed.). Springer.

Horst, A. M., Hill, A. P., & Gorman, K. B. (2020). *palmerpenguins: Palmer Archipelago (Antarctica) penguin data*. R package version 0.1.0. https://allisonhorst.github.io/palmerpenguins/

---

# Appendix: Additional Resources

## Useful Packages for Beginners

```{r eval=FALSE}
# Data manipulation
install.packages("dplyr")
install.packages("tidyr")

# Visualization
install.packages("ggplot2")

# Data import
install.packages("rio")
install.packages("readr")
install.packages("haven")       # For SPSS, Stata, SAS

# Reporting
install.packages("rmarkdown")
install.packages("knitr")

# Practice datasets
install.packages("palmerpenguins")
install.packages("datasets")
```

## Common Keyboard Shortcuts (RStudio)

- **Ctrl/Cmd + Enter**: Run current line/selection
- **Ctrl/Cmd + Shift + Enter**: Run entire script
- **Ctrl/Cmd + S**: Save file
- **Ctrl/Cmd + Shift + C**: Comment/uncomment
- **Alt/Option + -**: Insert `<-` assignment operator
- **Tab**: Auto-complete function/variable names

## Where to Get Help

1. **Built-in help**: `?function_name` or `help(function_name)`
2. **RStudio Community**: https://community.rstudio.com/
3. **Stack Overflow**: Tag questions with [r]
4. **R Documentation**: https://www.rdocumentation.org/
5. **Your instructor**: Office hours or course forum

## Tips for Success

1. **Practice regularly** - Programming is a skill learned by doing
2. **Type code yourself** - Don't just copy-paste
3. **Read error messages carefully** - They're trying to help!
4. **Use comments** - Explain your code with `#`
5. **Save often** - Use Ctrl/Cmd + S frequently
6. **Start simple** - Break complex problems into smaller steps
7. **Ask for help** - Everyone starts as a beginner!

Remember: R has a steep learning curve at first, but it gets easier with practice. You're building a valuable skill that will serve you throughout your analytics career!

---

*Document generated on `r format(Sys.Date(), '%B %d, %Y')`*
