---
title: "Week 2: Frameworks, Benchmarks, and Literature Reviews"
author: "Ziyuan Huang"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(DiagrammeR)
library(ggplot2)
library(ggrepel)
```

# Introduction

This week we move from a broad research idea to a structured, justifiable research plan. Good research is always **contextualized within the history of what has been done** — you must show that you understand the existing landscape before you can make a meaningful contribution to it. To do this, we focus on three foundational skills:

*   A **Framework** is the intellectual structure for your study — an existing analytic, theoretical, or methodological tradition that guides how you approach your problem and interpret your results.

*   A **Benchmark** is your yardstick for success — an established standard from prior work that lets you answer the question *"How good are my results compared to the best that's been done?"*

*   A **Literature Review** is the written argument that ties it all together — connecting existing research to your specific problem, establishing the importance of your research questions, and justifying your approach.

---

# Part 1: Frameworks

## What is a Framework?

For different types of problems, **frameworks** exist for the best — or at least most common — approaches to that class of problem. A framework is not something you invent; it is something you adopt, adapt, and justify. Your choice signals to readers that you understand the research tradition your work belongs to.

## Types of Frameworks

Research frameworks fall into three broad categories, and a single study may incorporate more than one:

```{r framework-types, fig.align='center'}
grViz("
digraph framework_types {
  graph [layout = dot, rankdir = LR, label = 'Three Types of Research Frameworks', labelloc = t, fontsize = 20, splines=ortho]

  node [shape = box, style = 'rounded,filled', fontname = Helvetica, fontsize = 13]
  edge [fontname = Helvetica]

  F [label = 'Research\nFramework', fillcolor = '#4A90D9', fontcolor = white, shape = ellipse]

  A [label = 'Analytic\nFramework', fillcolor = '#5BA85B', fontcolor = white]
  B [label = 'Theoretical\nFramework', fillcolor = '#D97B4A', fontcolor = white]
  C [label = 'Methodological\nFramework', fillcolor = '#9B59B6', fontcolor = white]

  A1 [label = 'Predictive Modeling', fillcolor = '#D5EDD5']
  A2 [label = 'Forecasting', fillcolor = '#D5EDD5']

  B1 [label = 'Game Theory', fillcolor = '#FDDFC8']
  B2 [label = 'Information Theory', fillcolor = '#FDDFC8']

  C1 [label = 'Experimental Design', fillcolor = '#E8D5F5']
  C2 [label = 'Data Mining', fillcolor = '#E8D5F5']

  F -> A; F -> B; F -> C;
  A -> A1; A -> A2;
  B -> B1; B -> B2;
  C -> C1; C -> C2;
}
")
```

### Analytic Frameworks

Analytic frameworks define the *type of analysis* you will conduct and the computational or statistical tools best suited for that analysis.

| Framework | Description | Example Use Case |
|---|---|---|
| Predictive Modeling | Build a model to estimate an unknown outcome from known inputs | Predicting student dropout from engagement data |
| Forecasting | Estimate future values of a time-series variable | Projecting enrollment trends over the next 5 years |
| Classification | Assign observations to discrete categories | Diagnosing disease from medical imaging |
| Clustering / Segmentation | Discover natural groupings in unlabeled data | Identifying patient risk profiles |

### Theoretical Frameworks

Theoretical frameworks provide the *conceptual lens* through which you interpret your problem and results. They connect your study to a body of scholarly thought.

| Framework | Description | Example Use Case |
|---|---|---|
| Game Theory | Models strategic decision-making among rational agents | Analyzing competitive pricing behavior |
| Information Theory | Quantifies information, uncertainty, and communication limits | Measuring entropy in communication networks |
| Tinto's Model | Explains student persistence through academic and social integration | Predicting college dropout |
| Technology Acceptance Model | Explains user adoption of new technologies | Studying e-learning platform adoption |

### Methodological Frameworks

Methodological frameworks define the *overall research design strategy* — how you will collect, generate, or analyze evidence.

| Framework | Description | Example Use Case |
|---|---|---|
| Experimental Design | Controlled manipulation of variables with randomization | Testing the effect of an intervention on outcomes |
| Quasi-Experimental | Comparison without full randomization | Pre/post analysis without a true control group |
| Data Mining | Extraction of patterns from large datasets | Mining electronic health records for adverse events |
| Survey Research | Systematic collection of self-reported data | Measuring student satisfaction across programs |

## Selecting and Justifying Your Framework

After reviewing the literature, choose the framework(s) that best fit your problem. A study may legitimately combine more than one — for example, using a **predictive modeling** analytic framework, guided by **Tinto's theoretical model**, and executed through a **data mining** methodological approach.

Your justification should explain:

1. **Why this framework fits your problem** — how its assumptions and constructs match your context.
2. **Any adaptations needed** — e.g., applying an on-campus student model to an online setting.
3. **How it connects to your research question** — the framework should make your RQ feel inevitable and well-motivated.

---

# Part 2: Benchmarks

## What is a Benchmark?

For the majority of analytic problems, **some model or solution already exists**. New models or solutions must be compared against them. As you define your problem and methodological approach, you must establish what the **baseline or best current performance** is for your task.

> Without benchmarks, your result is just a number without meaning. A model with 83% accuracy could be excellent or mediocre — it depends entirely on what the field has already achieved.

## Types of Benchmarks

```{r benchmark-types-diagram, fig.align='center'}
grViz("
digraph benchmark_types {
  graph [layout = dot, rankdir = TB, label = 'Types of Benchmarks', labelloc = t, fontsize = 20]

  node [shape = box, style = 'rounded,filled', fontname = Helvetica, fontsize = 13]

  B [label = 'Benchmarks', fillcolor = '#4A90D9', fontcolor = white, shape = ellipse]

  D [label = 'Benchmark\nDatasets', fillcolor = '#5BA85B', fontcolor = white]
  M [label = 'Benchmark\nMetrics', fillcolor = '#D97B4A', fontcolor = white]
  E [label = 'Exploratory /\nDescriptive', fillcolor = '#9B59B6', fontcolor = white]

  D1 [label = 'ImageNet\n(image classification)', fillcolor = '#D5EDD5']
  D2 [label = 'UCI Repository\n(general ML)', fillcolor = '#D5EDD5']
  D3 [label = 'GLUE\n(NLP tasks)', fillcolor = '#D5EDD5']

  M1 [label = 'Reported accuracy, F1,\nAUC for a specific task', fillcolor = '#FDDFC8']
  M2 [label = 'State-of-the-art\nperformance thresholds', fillcolor = '#FDDFC8']

  E1 [label = 'Compare to what is\ncurrently known about\nthe topic or area', fillcolor = '#E8D5F5']

  B -> D; B -> M; B -> E;
  D -> D1; D -> D2; D -> D3;
  M -> M1; M -> M2;
  E -> E1;
}
")
```

### Benchmark Datasets

Established datasets created specifically to test new models or solutions on common tasks. Using a standard benchmark dataset allows direct, apples-to-apples comparison with every other study that used it.

**Examples:**

*   **ImageNet** — the standard for basic image classification (millions of labeled images across 1,000 classes)
*   **UCI Machine Learning Repository** — a broad collection of datasets across domains, widely used in academic research
*   **GLUE / SuperGLUE** — standard benchmarks for natural language understanding tasks

### Benchmark Metrics

Recorded performance values from prior high-quality studies for a given task under given conditions. These answer the question: *"What is the best performance anyone has achieved on this specific problem?"*

When reporting benchmark metrics, always capture:

| Field | Description |
|---|---|
| Author & Year | Who achieved this result |
| Task / Problem | Exactly what problem was being solved |
| Method | What model or approach was used |
| Performance Metric | The metric used (Accuracy, F1, AUC, RMSE, etc.) |
| Result | The numeric value |
| Conditions | Dataset, data splits, key parameters |

### Exploratory and Descriptive Research Benchmarks

Exploratory or descriptive research generally does not have concrete numeric benchmarks, but it must still be compared to **what is currently known** about the topic or area. Your contribution is measured by how much new understanding you add relative to the existing body of knowledge — what patterns were previously unknown, what relationships were previously assumed but not demonstrated, or what populations were previously understudied.

## Visualizing Your Result Against Benchmarks

```{r benchmark-chart, fig.align='center'}
benchmark_data <- data.frame(
  study = c('Smith et al. (2019)\n(Baseline)', 'Chen (2020)', 'Jones & Lee (2021)\n(State-of-the-Art)', 'Your Hypothetical\nResult'),
  f1_score = c(0.78, 0.82, 0.85, 0.83),
  type = c('Baseline', 'Prior Work', 'State-of-the-Art', 'Your Result')
)

ggplot(benchmark_data, aes(x = reorder(study, f1_score), y = f1_score, fill = type)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = f1_score), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c('Baseline' = '#d9534f',
                               'Prior Work' = '#5bc0de',
                               'State-of-the-Art' = '#5cb85c',
                               'Your Result' = '#f0ad4e')) +
  labs(
    title = 'Visualizing Your Result Against Benchmarks',
    subtitle = 'Context makes your performance meaningful',
    x = 'Study',
    y = 'F1-Score',
    fill = 'Study Type'
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  ylim(0, 1.0)
```

**Reading the chart:** Your hypothetical F1 of 0.83 sits above the baseline (0.78) and prior work (0.82), and is competitive with the SOTA (0.85). Without this comparison, 0.83 has no interpretable meaning.

---

# Part 3: Establishing Frameworks and Benchmarks via a Literature Search

## How to Conduct a Literature Search

To determine the appropriate frameworks and benchmarks for your study, you must conduct a **literature search**. The goal is to accumulate 10–20 peer-reviewed sources, after which you have enough foundation to begin writing your literature review section.

### Where to Search

| Source | Notes |
|---|---|
| **Google Scholar** | Best starting point; broad coverage; shows citation counts |
| **Ask ChatGPT** | Can suggest frameworks and papers — but **always verify the sources exist and say what is claimed** |
| **Reference sections** | Once you find a key paper, mine its reference list for related foundational work |

> **Source quality:** Peer-reviewed journal articles and conference papers are the gold standard. White papers from reputable institutions are acceptable. Blog posts, news articles, and other informal sources are **not** acceptable as primary citations.

### Building Your Search Query

```{r keyword-search-viz, fig.align='center'}
grViz("
digraph keyword_search {
  graph [layout = dot, rankdir = LR, labelloc = t, fontsize = 18, fontname = Helvetica, splines = ortho]

  node [shape = box, style = 'rounded,filled', fontname = Helvetica, fontsize = 12, fillcolor = white]

  subgraph cluster_topic {
    label = '1. Start with Topic';
    style = filled; color = '#ECECEC';
    topic [label = 'Predicting Student Dropout\nin Online Courses', shape = box, style = 'rounded,filled', fillcolor = '#D6E4F0', fontcolor = '#1A1A1A'];
  }

  subgraph cluster_concepts {
    label = '2. Core Concepts';
    style = filled; color = '#D6EAF8';
    c1 [label = 'Student Dropout',   shape = box, style = 'rounded,filled', fillcolor = '#2E86C1', fontcolor = white];
    c2 [label = 'Online Courses',    shape = box, style = 'rounded,filled', fillcolor = '#2E86C1', fontcolor = white];
    c3 [label = 'Machine Learning',  shape = box, style = 'rounded,filled', fillcolor = '#2E86C1', fontcolor = white];
  }

  subgraph cluster_synonyms {
    label = '3. Expand with Synonyms (OR)';
    style = filled; color = '#D5F5E3';
    s1 [label = 'attrition\npersistence',            shape = box, style = 'rounded,filled', fillcolor = '#1E8449', fontcolor = white];
    s2 [label = 'e-learning\nMOOCs',                 shape = box, style = 'rounded,filled', fillcolor = '#1E8449', fontcolor = white];
    s3 [label = 'predictive modeling\ndata mining',  shape = box, style = 'rounded,filled', fillcolor = '#1E8449', fontcolor = white];
  }

  subgraph cluster_framework_terms {
    label = '4. Add Framework Terms (AND)';
    style = filled; color = '#FDEBD0';
    f1 [label = 'theory\nframework\nreview', shape = box, style = 'rounded,filled', fillcolor = '#CA6F1E', fontcolor = white];
  }

  subgraph cluster_query {
    label = '5. Final Search Query';
    style = filled; color = '#922B21';
    query [label = 'student dropout OR attrition\nAND online course OR e-learning\nAND theory OR framework OR review', shape = box, style = 'rounded,filled', fillcolor = '#922B21', fontcolor = white, color = white, fontsize = 13];
  }

  topic -> {c1, c2, c3}
  c1 -> s1; c2 -> s2; c3 -> s3;
  {s1, s2, s3, f1} -> query;
}
")
```

## Benchmark Literature Table

As you accumulate sources, track benchmark studies in a structured table:

| Author (Year) | Task / Problem | Method | Metric | Result | Notes |
|---|---|---|---|---|---|
| Smith et al. (2019) | Predicting dropout in MOOCs | Logistic Regression | Accuracy | 78% | Demographic features only. Good baseline. |
| Jones & Lee (2021) | Real-time dropout prediction | Random Forest | F1-Score | 0.85 | Clickstream data. Considered SOTA. |
| Chen (2020) | Identifying at-risk students | SVM | AUC | 0.82 | First two weeks of course only. |

**Synthesis statement template:** *"The current literature shows that for [your task], state-of-the-art models achieve [SOTA metric]. Simpler baseline models achieve [baseline metric]. My goal is to exceed this baseline while [your specific contribution — e.g., using a more interpretable model, applying to a novel population, etc.]."*

---

# Part 4: The Literature Review

## Purpose of a Literature Review

A literature review is not a summary exercise — it is a **structured argument** that does three specific things:

1. **Relates your work to ongoing discussion in the field** — situating your project within the broader scholarly conversation.
2. **Establishes a framework for the importance of your research question** — demonstrating why your question matters and why it has not been fully answered.
3. **Establishes benchmarks for comparing results and conclusions** — answering "How much better is your model?", "How much more variance can you explain?", and "What have you added to what is known?"

## What a Literature Review IS

A well-written literature review is:

*   **A logical argument** based on a comprehensive understanding of the state of research on your topic
*   **Fully supported by credible sources** — every major claim cites peer-reviewed evidence
*   **Explicitly connective** — it shows how your proposed project relates to past work, and very importantly, how your proposed work **extends past studies or addresses their limitations**
*   **Organized with logical flow** — it moves coherently from the broad context of your field to your specific research questions

```{r lit-review-anatomy, fig.align='center'}
grViz("
digraph lit_review {
  graph [layout = dot, rankdir = TB, label = 'Anatomy of a Literature Review Argument', labelloc = t, fontsize = 18]

  node [shape = box, style = 'rounded,filled', fontname = Helvetica, fontsize = 12]

  A [label = 'Broad Field Context\n(What is the general problem area?)', fillcolor = '#AED6F1']
  B [label = 'Existing Approaches & Frameworks\n(What has been tried? What theories exist?)', fillcolor = '#A9DFBF']
  C [label = 'Gaps & Limitations in Prior Work\n(What has NOT been answered well?)', fillcolor = '#FAD7A0']
  D [label = 'Your Research Question\n(How does your study address a gap?)', fillcolor = '#F1948A']
  E [label = 'Your Hypotheses / Expected Contributions\n(What will you show or demonstrate?)', fillcolor = '#D7BDE2']

  A -> B [label = ' grounds ']
  B -> C [label = ' reveals ']
  C -> D [label = ' motivates ']
  D -> E [label = ' leads to ']
}
")
```

## What a Literature Review is NOT

Avoid these common mistakes:

| What students write | What it actually is | Why it fails |
|---|---|---|
| "Smith (2020) found that X. Jones (2021) found that Y. Chen (2022) found that Z." | An annotated bibliography | No argument, no connection, no synthesis |
| A list of paper summaries with no linking commentary | A data dump | Reader cannot see how it all relates to your project |
| Extensive technical explanation of algorithms you will use | A methods primer | Unless you are developing a new algorithm, briefly justify *why* the method fits — don't explain how it works |

> A good literature review reads like a coherent essay that happens to have citations, not like a collection of summaries stapled together.

## Citation Management with Zotero

Managing 10–20+ sources manually is error-prone. **Zotero** (https://www.zotero.org/) is the recommended free, open-source citation management tool.

**What Zotero does:**

*   Stores full citation metadata (author, year, journal, DOI, etc.)
*   Generates formatted citations in APA, MLA, Chicago, and other styles automatically
*   Lets you attach PDFs, highlight passages, and keep notes on each source
*   Syncs across devices and integrates with Word, Google Docs, and RStudio

HU's Library has put together resources and guides to help you get started with Zotero.

---

# Part 5: Organizing Your Literature Review

## Three Methods for Outlining a Literature Review

Before writing, you need a plan. Three proven methods for organizing a literature review are:

### Method 1: Literature Maps

A **literature map** is a diagram that links important concepts and clusters your sources around those concepts. It is the most visual approach and is excellent for seeing the structure of a large body of literature at a glance.

```{r lit-map, fig.align='center'}
grViz("
digraph lit_map {
  graph [layout = dot, rankdir = TB, label = 'Example Literature Map: Twitter Marketing & Tech Stocks', labelloc = t, fontsize = 16, splines=curved]

  node [shape = ellipse, style = 'filled', fontname = Helvetica, fontsize = 11]
  edge [fontname = Helvetica, fontsize = 10]

  RQ [label = 'Research Question:\nInfluence of Twitter Marketing\non Tech Stocks', fillcolor = '#4A90D9', fontcolor = white, shape = box, style = 'rounded,filled']

  TM [label = 'Twitter Marketing', fillcolor = '#5BA85B', fontcolor = white]
  TI [label = 'Tech Industry Marketing', fillcolor = '#D97B4A', fontcolor = white]
  TS [label = 'Factors Impacting\nTech Stocks', fillcolor = '#9B59B6', fontcolor = white]

  TM1 [label = 'Taecharungroj (2017)\nStarbucks Twitter strategy', fillcolor = '#D5EDD5', shape = note]
  TM2 [label = 'Leung et al. (2015)\nFacebook vs. Twitter\nhotel industry', fillcolor = '#D5EDD5', shape = note]

  TI1 [label = 'Nouri & Ahmady (2018)\nMarketing in high-tech SMBs', fillcolor = '#FDDFC8', shape = note]
  TI2 [label = 'Yang & Gabrielsson (2017)\nEntrepreneurial marketing B2B', fillcolor = '#FDDFC8', shape = note]

  TS1 [label = 'Bento et al. (2019)\nCrowdfunding & clean-tech stocks', fillcolor = '#E8D5F5', shape = note]
  TS2 [label = 'Lee et al. (2013)\nIndustry returns & stock market', fillcolor = '#E8D5F5', shape = note]

  RQ -> TM; RQ -> TI; RQ -> TS;
  TM -> TM1; TM -> TM2;
  TI -> TI1; TI -> TI2;
  TS -> TS1; TS -> TS2;
}
")
```

Many templates and examples for literature maps are available online. Search for *"literature map examples template"* to find formats that match your project's complexity.

### Method 2: Subheadings

Essentially a literature map formatted as a **bulleted outline**. Each subheading represents a major concept or theme, and sources are listed under the subheading they belong to.

**Rules of thumb:**
*   Start from the **broadest concepts and oldest, most foundational methods** and progressively narrow to what is unknown or evolving
*   No more than **2 subheadings per page** of finished text
*   At least **4 citations under each subheading**

**Example Topic:** *The influence of Twitter marketing on tech stocks*

---

**Subheading 1: Twitter Marketing**

- Taecharungroj, V. (2017). Starbucks' marketing communications strategy on Twitter. *Journal of Marketing Communications, 23*(6), 552–571.
- Leung, X. Y., Bai, B., & Stahura, K. A. (2015). The marketing effectiveness of social media in the hotel industry: A comparison of Facebook and Twitter. *Journal of Hospitality & Tourism Research, 39*(2), 147–169.

**Subheading 2: Tech Industry Marketing**

- Nouri, P., & Ahmady, A. (2018). A taxonomy of nascent entrepreneurs' marketing decisions in high-tech small businesses. *Journal of Small Business Strategy, 28*(3), 69–79.
- Yang, M., & Gabrielsson, P. (2017). Entrepreneurial marketing of international high-tech business-to-business new ventures. *Industrial Marketing Management, 64*, 147–160.

**Subheading 3: Factors Impacting Tech Stocks**

- Bento, N., Gianfrate, G., & Groppo, S. V. (2019). Do crowdfunding returns reward risk? *Technological Forecasting and Social Change, 141*, 107–116.
- Lee, C. C., Chen, M. P., & Chang, C. H. (2013). Dynamic relationships between industry returns and stock market returns. *The North American Journal of Economics and Finance, 26*, 119–144.

---

### Method 3: First Sentences

The most rigorous organizational method. Instead of just listing topics, you write **the actual first sentence of each paragraph** in your literature review before writing the full text. Each first sentence is a claim or premise that your supporting citations will back up.

This approach forces you to think through the **logical argument** your literature review is making — every sentence must connect to the next, moving from broad context toward your specific research questions.

**Example Topic:** *Changes in communication styles of political leaders*

> 1. Most scholars and political commentators agree that Donald Trump is unlike any previous US president.
> 2. During his campaign and early presidency, many claimed that Trump's simplicity and directness were keys to his popularity.
> 3. A second major distinguishing feature of Trump's language is his self-confidence.
> 4. We build on an extensive research tradition of exploring political leadership traits.
> 5. If conveying simple messages with confidence is a powerful way to persuade voters, it is incumbent on researchers to determine effective ways to measure these dimensions reliably, quickly, and efficiently.
> 6. A common distinction in language is between content words and function words.
> 7. The eight categories of function words capture an underlying psychological dimension known as analytic thinking.
> 8. Past work has validated the use of function words as markers of the analytic thinking dimension, including a study of college admission essays from over 25,000 incoming students.
> 9. The analysis of function words has also been useful in identifying people's relative status or "clout" in a social hierarchy.
> 10. The computerized text analysis program LIWC was updated in 2015 to include empirically established measures of both analytic thinking and confidence.
> 11. The algorithm developed from Kacewicz et al. is formally referred to as *clout*.
> 12. The purpose of this project was to apply linguistic analyses to large corpora of texts from the American presidency, non-US leaders, legislative bodies, and cultural contexts spanning multiple decades.

Notice how each sentence builds logically on the last — this is a **complete argumentative skeleton** for a literature review before a single full paragraph is written.

---

# Part 6: Final Guidance on Literature Reviews

```{r final-guidance-viz, fig.align='center'}
grViz("
digraph guidance {
  graph [layout = dot, rankdir = LR, label = 'Do vs. Do Not', labelloc = t, fontsize = 18, splines=ortho]

  node [fontname = Helvetica, fontsize = 12]

  subgraph cluster_do {
    label = 'DO';
    style = filled; color = '#D5EDD5';
    node [style='filled', fillcolor=white, shape=box];
    d1 [label = 'Read critically']
    d2 [label = 'Write short summaries\nfor each source']
    d3 [label = 'Keep your goal in mind\nthroughout']
    d4 [label = 'Find multiple citations\nfor main points']
    d5 [label = 'Focus on the big picture']
    d6 [label = 'Briefly justify WHY\na method fits your problem']
  }

  subgraph cluster_dont {
    label = 'DO NOT';
    style = filled; color = '#FDDFC8';
    node [style='filled', fillcolor=white, shape=box];
    n1 [label = 'Just summarize sources\none by one']
    n2 [label = 'Write an\nannotated bibliography']
    n3 [label = 'Dump every fact\nyou can find']
    n4 [label = 'Extensively explain\ntechnical components']
    n5 [label = 'Lose sight of\nyour research question']
    n6 [label = 'Use informal sources\n(blogs, news, etc.)']
  }
}
")
```

**Key reminders:**

*   Read critically — ask yourself *"Does this source support my argument? Does it contradict it? Does it reveal a gap?"*
*   Write short notes or summaries as you read each source — you will not remember the details later
*   Keep your research goal visible while writing — every paragraph should connect back to it
*   Try to find **multiple citations** supporting your main points — a claim backed by 3–4 sources is much stronger than one backed by 1
*   Focus on the **big picture**, not small technical details — unless you are developing a new algorithm, you do not need to extensively explain technical components; briefly discuss why the chosen model makes the most sense for your particular problem

---

# Bringing It All Together

Your framework, benchmarks, and literature review are not separate deliverables — they are one integrated argument.

```{r research-narrative-viz, fig.align='center'}
grViz("
digraph research_narrative {
  graph [layout = dot, rankdir = TB, label = 'From Literature to Contribution', labelloc = t, fontsize = 18]

  node [shape = box, style = 'rounded,filled', fontname = Helvetica, fontsize = 12]

  A [label = 'Literature Search\n(10-20 peer-reviewed sources)', fillcolor = '#AED6F1']
  B [label = 'Adopt a Framework\n(Analytic / Theoretical / Methodological)', fillcolor = '#A9DFBF']
  C [label = 'Establish Benchmarks\n(Datasets / Metrics / Known patterns)', fillcolor = '#FAD7A0']
  D [label = 'Write Literature Review\n(Logical argument connecting field to your RQ)', fillcolor = '#D7BDE2']
  E [label = 'Design Methodology\n(Guided by framework and gap identified)', fillcolor = '#F9E79F']
  F [label = 'Generate & Compare Results\n(Against established benchmarks)', fillcolor = '#FADBD8']
  G [label = 'State Your Contribution\n(What you added to the field)', fillcolor = '#D5EDD5']

  A -> B; A -> C; A -> D;
  B -> D; C -> D;
  D -> E;
  E -> F;
  F -> G;
}
")
```

**Example narrative statement:**

> "My research uses a predictive modeling analytic framework, grounded in an adapted version of Tinto's Student Integration Model, to predict student dropout in online courses. A literature search revealed that the state-of-the-art F1-Score for this task is 0.85 (Jones & Lee, 2021), using clickstream features, with a baseline of 78% accuracy from simpler demographic models (Smith et al., 2019). My study extends this work by applying theoretically-grounded features tied explicitly to academic and social integration constructs, which prior benchmark studies did not use. By doing so, my model aims to be both predictive and interpretable — addressing a limitation in existing SOTA work that prioritizes performance over explainability."

---

# Next Steps & To-Do

- [ ] **Action 1:** Identify which type(s) of framework apply to your project (Analytic, Theoretical, Methodological) and locate the foundational paper for your chosen framework.
- [ ] **Action 2:** Conduct a literature search on Google Scholar. Collect 10–20 peer-reviewed sources and save them in Zotero.
- [ ] **Action 3:** Identify 2–3 benchmark studies and record their task, method, metric, and result in a table.
- [ ] **Action 4:** Choose an organizational method (Literature Map, Subheadings, or First Sentences) and draft an outline for your literature review before writing full text.
- [ ] **Action 5:** Write a 1-paragraph synthesis statement connecting your research question, your framework, and the key benchmarks you will use for comparison.
